Project Description -- Erosion, Transport, Anomalous Diffusion, and Collapse in Granular and Porous Media

### 2019-11-16 docs/proposals/NSF/ProjDesc/ProjDesc.tex
worked on some of the broader impact stuff

\begin{figure}[htp]
  \begin{center}
  \begin{subfigure}[b]{0.45\textwidth}
  \includegraphics[width=\textwidth]{figs/tort_eulerian100}
  \caption{100 eroding bodies}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\textwidth}
  \includegraphics[width=\textwidth]{figs/tort_eulerian20}
  \caption{20 eroding bodies}
  \end{subfigure}
  \end{center}
  \caption{\label{fig:100tortuosity} The tortuosity of two eroding geometries. The tortuosity can be computed using two different formulas (red vs.~blue marks). In the left plot, the red square is the tortuosity of the geometry in Figure~\ref{fig:100tracers}. In both cases, there is qualitative agreement between our numerical simulations and a power-law model (black dashed line)~\cite{mat-kha-koz2008}.}
\end{figure}

@@

One central goal of this proposal is to enable the free motion of grains during the erosion processes. As mentioned above, this goal requires resolving contact forces between bodies, as well as the incorporation of the external buoyancy forces. Once this goal is achieved, it will be possible to simulate a complex porous-medium of arbitrarily sized and shaped grains undergoing the simultaneous action of mechanical-or-chemical erosion and transport due to the combined effects of seepage, contact, and buoyancy forces. These computations will thus simulate, with high-fidelity, the multi-physics of realistic porous-medium dynamics, albeit in a scaled down context. Ultimately, such simulations could be used to build an understanding of medium reconfiguration events, such as sinkhole collapse or conduit formation in karst networks.

	In very preliminary work, PI Moore has been running simplified simulations with a current graduate student. These preliminary simulations are based on a hybrid continuum-discrete model for the permeability field and grain particle field respectively. They utilize the homogenized description of the flow field in \eqref{DarcyEq} and incorporate the physical effects of:
\begin{itemize}[noitemsep]
\item Transport of grains due to seepage force from the surrounding fluid flow.
\item Gravitational sedimentation of grains.
\item Feedback of grain-distribution onto permeability field.
\item Contact and cohesive forces between grains.
\end{itemize}
Preliminary investigation suggests that this last mechanisms of cohesion is essential to model sinkhole collapse in the system.

These simulation have not yet begun to incorporate the slow-acting effects of mechanical and chemical erosion, and the associated gradual changes in medium properties. A grand goal of this proposal would be to merge the two lines of inquiry to develop a sound theoretical framework for porous-medium dynamics robust enough to capture both short term events, such as sinkhole collapse, and long-term changes in the medium. The greatest challenge faced is the disparate temporal and spatial scales that must be considered. Our proposed strategy is to use the high-fidelity erosion simulations to parametrize the effect of erosion on medium properties.

	More specifically, we plan to run large batches of the high-fidelity boundary-integral-equation-based numerical simulations of erosion and then leverage rapid advances in the field of machine learning and deep-neural networks to develop reduced models that parametrize the effects of erosion on the evolution of macroscopic medium properties. The output of the high-fidelity simulations can be transformed into essential an image format, enabling us to use cutting-edge advancements on deep neural-networks for image analysis \cite{pelt2018mixed}. Some important advantages of referenced work are the use of dilated convolutions in a deeply connected network, which first mixes scales and second reduces the number of super-parameters present and thus largely averts the risk of overfitting data. Both PIs will supervise a graduate student who will design and implement the deep-neural network for learning the essence of the high-fidelity simulations. Once the essential features are learned, they will be incorporated into reduced-order models that utilize the multiphase framework to enable more efficient and larger scale coarse-grained simulations of dynamic erosion.
        \todo[inline]{Nathan Kutz has a 2017 JFM 4 page paper that talks
        about neural networks/deep learning in fluid dynamics}

As mentioned previously, researchers at FSU's Geophysical Fluid Dynamics Institute (GFDI) have investigated the mechanisms behind sinkhole formation with a series of scaled-down laboratory experiments~\cite{tao2014experimental}. Firstly, already existing data from these experiments offers a testbed for comparison against our newly developed simulations. Furthermore, we propose to run a set of new experiments at the facilities offered by GFDI, spearheaded by an undergraduate student in close collaboration with the graduate students and PIs. These new experiments will more deeply probe the mechanisms underlying sinkhole formation with the special design goal of comparison agains the numerical simulations. These experiments will be used to calibrate and validate our computations.

@@

	This research proposed here could lead to a range of societal benefits, including better management of water resources, accurate prediction of contaminant transport, and an understanding of the long-term effects of human activities, such as hydraulic fracturing or groundwater pumping, on porous media and water resources. In such scenarios, the slow timescales of mechanical and chemical erosion are coupled to the faster timescales of human-induced changes and gravitational collapse. Currently, the relationship between these scales is poorly understood. The high-fidelity, efficient computational methods proposed here will enable a deeper understanding of these processes, which is the first step towards developing effective strategies for managing water resources and mitigating catastrophic events such as contamination or structural collapse. For example, the insights gained here could potentially help identify locations in natural aquifers vulnerable to contamination and/or collapse.

@@

\subsection{Graduate research}

	An important benefit of the research proposed here is the potential to immerse graduate students in fundamentally {\em cross-disciplinary} research. This proposal involves a confluence of disciplines, most centrally computational mathematics, but also geophysics, reduced-physical modeling, hydrology, and environmental science. Exposure to a range of disciplines is, in the view of the PIs, an invaluable opportunity for young scientists.

	Naturally, the research proposed here will form the foundation for the PhD dissertations of the graduate students involved. Beyond this, the students will engage in publishing articles in peer-reviewed journals of high quality and attending national and international conferences at which they will contribute presentations on their work. The PIs will closely mentor the graduate students to ensure their continued development, not only intellectually, but also in the `soft' skills of communication, promotion, and networking. Having both PIs Moore and Quaife involved in the training of each graduate student is especially valuable for offering multiple perspectives on these issues. The PIs will make special efforts to recruit women and underrepresented minorities as the graduate students funded by this project. PI Moore has already had success recruiting women in these roles, as the first graduate student to receive her PhD under Moore's advising was female (Karina Khazmutdinova, 2016).

\subsection{Undergraduate research}

Regarding undergraduate research, the ideas in this proposal represent a rich and fascinating topic for undergraduate students, while also being approachable these students. In particular, we aim to recruit at least one undergraduate student to perform laboratory experiments at the Geophysical Fluid Dynamics Institute (GFDI). As mentioned above, GFDI offers state-of-the-art laboratory facilities, with an already-functioning sinkhole experiment. The student will perform controlled experiments using natural and artificial porous media (e.g.~sand and micro-glass beads respectively) to precisely characterize the parameter regime leading to gravitational collapse. Parallel experiments will be conducted with erodible (bentonite clay) or dissolvable  (sucrose solids) fragments materials, which will represent the porous medium. These experiments will be used to guide, calibrate, and validate the computational advances. These undergraduate students will work closely with the graduate students to ensure that theory and experiments advance in close cooperation, with constant feedback between the two.

@@

\subsection{Outreach for K-12 and the general public}

The PIs have already seized the opportunity to engage K-12 students in outreach activities. In particular, Moore's research was the focus of an educational video created by {\bf CPALMS}, which is the State of Florida's official source for standards information and course descriptions for K-12 education. The video, which features an interview and animations of research, is used to reinforce concepts from mathematics courses (grade levels 7-12), and to encourage students to consider a career in the STEM fields~\cite{CPALMS}. Additionally, these videos are frequently used for the continuing education of K-12 {\em teachers}, providing them a broader perspective of mathematics that could be integrated into their teaching.

>>>>

\begin{figure}[htp]
  \begin{center}
  \begin{subfigure}[b]{0.45\textwidth}
  \includegraphics[width=\textwidth]{figs/tort_eulerian100}
  \caption{100 eroding bodies}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\textwidth}
  \includegraphics[width=\textwidth]{figs/tort_eulerian20}
  \caption{20 eroding bodies}
  \end{subfigure}
  \end{center}
  \caption{\label{fig:100tortuosity} The tortuosity of two different eroding geometries. The tortuosity can be computed using two different formulas (red vs.~blue marks). In the left plot, the red square is the tortuosity of the geometry in Figure~\ref{fig:100tracers}. In both cases, there is qualitative agreement between our numerical simulations and a power-law model (black dashed line)~\cite{mat-kha-koz2008}.}
\end{figure}

@@

One central goal of this proposal is to enable the free motion of grains during the erosion processes. As mentioned earlier, this goal requires resolving contact forces between bodies, as well as the incorporation of the external buoyancy forces. Once this goal is achieved, it will be possible to simulate a complex porous-medium of arbitrarily sized and shaped grains undergoing the simultaneous action of mechanical-or-chemical erosion and transport due to the combined effects of seepage, contact, and buoyancy forces. These computations will thus simulate, with high-fidelity, the multi-physics of realistic porous-medium dynamics, albeit in a scaled down context. Ultimately, such simulations could be used to build an understanding of medium reconfiguration events, such as sinkhole collapse or conduit formation in karst networks.

In very preliminary work, PI Moore has been running simplified simulations with a current graduate student. These preliminary simulations are based on a hybrid continuum-discrete model for the permeability field and grain particle field respectively. They utilize the homogenized description of the flow field in~\eqref{DarcyEq} and incorporate the physical effects of:
\begin{itemize}[noitemsep]
\item Transport of grains due to seepage force from the surrounding fluid flow.
\item Gravitational sedimentation of grains.
\item Feedback of grain-distribution onto permeability field.
\item Contact and cohesive forces between grains.
\end{itemize}
Preliminary investigation suggests that this last mechanisms of cohesion is essential to model sinkhole collapse in the system.

These simulation have not yet begun to incorporate the slow-acting effects of mechanical and chemical erosion, and the associated gradual changes in medium properties. A grand goal of this proposal would be to merge the two lines of inquiry to develop a sound theoretical framework for porous-medium dynamics robust enough to capture both short-term events, such as sinkhole collapse, and long-term changes in the medium. The greatest challenge faced is the disparate temporal and spatial scales that must be considered. Our proposed strategy is to use the high-fidelity erosion simulations to parametrize the effect of erosion on medium properties.

More specifically, we plan to run large batches of the high-fidelity boundary-integral-equation-based numerical simulations of erosion and then leverage rapid advances in the field of machine learning and deep-neural networks to develop reduced models that parametrize the effects of erosion on the evolution of macroscopic medium properties and the flow.  The output of the high-fidelity simulations of the grain configuration can be transformed into essentially an image format, enabling us to use cutting-edge advancements on deep neural-networks for image analysis~\cite{pelt2018mixed}. Some important advantages of referenced work are the use of dilated convolutions in a deeply connected network, which first mixes scales and second reduces the number of super-parameters present and thus largely averts the risk of overfitting data. Additionally, high-fidelity flow conditions can be used to train a neural network, and this has been successfully used by others to predict, as one example, an anisotropy tensor~\cite{ling2016reynolds}.  Both PIs will supervise a graduate student who will design and implement the deep-neural network for learning the essence of the high-fidelity simulations. Once the essential features are learned, they will be incorporated into reduced-order models that utilize the multiphase framework to enable more efficient and larger scale coarse-grained simulations of dynamic erosion.

As mentioned previously, researchers at FSU's Geophysical Fluid Dynamics Institute (GFDI) have investigated the mechanisms behind sinkhole formation with a series of scaled-down laboratory experiments~\cite{tao2014experimental}. First, already existing data from these experiments offers a testbed for comparison against our newly developed simulations. Furthermore, we propose to run a set of new experiments at the facilities offered by GFDI, spearheaded by an undergraduate student in close collaboration with the graduate students, PIs, and an experimentalist at GFDI. These new experiments will more deeply probe the mechanisms underlying sinkhole formation with the special design goal of comparison against the numerical simulations. These experiments will be used to calibrate and validate our computations.

@@

The proposed research could lead to a range of societal benefits, including better management of water resources, accurate prediction of contaminant transport, and an understanding of the long-term effects of human activities, such as hydraulic fracturing or groundwater pumping, on porous media and water resources. In such scenarios, the slow timescales of mechanical and chemical erosion are coupled to the faster timescales of human-induced changes and gravitational collapse. Currently, the relationship between these scales is poorly understood. The high-fidelity, efficient computational methods proposed here will enable a deeper understanding of these processes, which is the first step towards developing effective strategies for managing water resources and mitigating catastrophic events such as contamination or structural collapse. For example, the insights gained here could potentially help identify locations in natural aquifers vulnerable to contamination and/or collapse.

@@

\subsection{Graduate research}
An important benefit of the research proposed here is the potential to immerse graduate students in fundamentally {\em cross-disciplinary} research. This proposal involves a confluence of disciplines, most centrally computational mathematics, but also geophysics, reduced-physical modeling, hydrology, and environmental science.  Exposure to a range of disciplines is, in the view of the PIs, an invaluable opportunity for young scientists.

Naturally, the research proposed here will form the foundation for the
PhD dissertations of the graduate students involved. Beyond this, the
students will engage in publishing articles in peer-reviewed journals of
high quality and attending national and international conferences where
they will present their work and network with experts in the field. The
PIs will closely mentor the graduate students to ensure their continued
development, not only intellectually, but also in the `soft' skills of
communication, promotion, and networking. Having both PIs Moore and
Quaife involved in the training of each graduate student is especially
valuable for offering multiple perspectives on these issues. The PIs
will make special efforts to recruit women and underrepresented
minorities as the graduate students funded by this project. Both PIs
have already had success recruiting women in these roles, as the first
graduate student to receive her PhD under Moore's advising was female
(Karina Khazmutdinova, 2016), and Quaife's current PhD student is female
(Ashley Gannon, 2022).

\todo[inline]{Bryan is here}
\subsection{Undergraduate research}
Regarding undergraduate research, the ideas in this proposal represent a rich and fascinating topic for undergraduate students, while also being approachable these students. In particular, we aim to recruit at least one undergraduate student to perform laboratory experiments at the Geophysical Fluid Dynamics Institute (GFDI). As mentioned above, GFDI offers state-of-the-art laboratory facilities, with an already-functioning sinkhole experiment. The student will perform controlled experiments using natural and artificial porous media (e.g.~sand and micro-glass beads respectively) to precisely characterize the parameter regime leading to gravitational collapse. Parallel experiments will be conducted with erodible (bentonite clay) or dissolvable  (sucrose solids) fragments materials, which will represent the porous medium. These experiments will be used to guide, calibrate, and validate the computational advances. These undergraduate students will work closely with the graduate students to ensure that theory and experiments advance in close cooperation, with constant feedback between the two.

@@

\subsection{Outreach for K--12 and the general public}

\todo[inline]{FSU's YSP}
The PIs have already seized the opportunity to engage K-12 students in outreach activities. In particular, Moore's research was the focus of an educational video created by {\bf CPALMS}, which is the State of Florida's official source for standards information and course descriptions for K-12 education. The video, which features an interview and animations of research, is used to reinforce concepts from mathematics courses (grade levels 7-12), and to encourage students to consider a career in the STEM fields~\cite{CPALMS}. Additionally, these videos are frequently used for the continuing education of K-12 {\em teachers}, providing them a broader perspective of mathematics that could be integrated into their teaching.
